Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-05-23 doc/manuscript.Rmd
Updated from Duhaime suggestions.

***Journal***: Genome Research ([*Preparation Details*](http://genome.cshlp.org/site/misc/ifora_mspreparation.xhtml))

@@

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying microbes and viruses independently <!--I'd changed this previously, as "in isolation" can conjucture a whole slew of other things -->, thereby reducing them to two separate communities. Such approaches are unable to capture how these microbial communities interact, such as through processes that maintain community stability or allow phage-host populations to co-evolve. We developed and implemented a network-based analytical approach to describe phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages could infect which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. This study quantified and contrasted the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to better understand system perturbations, such as disease states, through ecological networks.

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages) have been also <!-- that was a split infinitive, you can't do that, also has to come after have been --> associated with various diseases and perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. The viruses act in concert with their microbial hosts as a single ecological community [@Haerter:2014ii]. Viruses influence their living microbial host communities through processes including lysis, modulating host gene expression [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja], influencing evolutionary processes, such as horizontal gene transfer [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn] or antagonistic co-evolution [@Koskella:2014ds], and by altering ecosystem processes and elemental stoichiometry [@Jover:2014gq].

Previous human microbiome work has focused on bacterial and viral communities, but have reduced them to two separate communities by studying them independently  [@Norman:2015kb; @Manrique:2016dx; @Ly:2014ew; @Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of bacteria or phage can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. Relationship-based network approaches allow us to capture this interaction information. We leveraged machine learning and graph theory techniques to characterize human bacterial and phage communities by their interferred relationships. <!--this is redundant to the previous sentence and distracting...not a good end of this paragraph-->

We characterized human skin and gut virus-bacteria network diversity to provide a foundation for further studies of disease network dynamics and to gain broader insights into human microbiome network diversity across different body sites.

Human bacterial-phage network diversity was established and explored by utilizing three published microbiome datasets that contained paired virus and whole community metagenomic sequence sets dominated by bacteria [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models [@Edwards:2015iz, <!--Lima-Mendez et al., Science 2015; her Tara paper, they predicted phage-host interactions and described phage-host ecology/host range distributions and compared to the Flores work of nestedness/modularity of host ranges-->, @Roux:2016cc] by its inclusion of protein interaction data, as well as experimentally verified negative and positive interactions, i.e., "who does and does not infect whom." Through this approach, we were able to provide a basic understanding of the network dynamics associated with phage and bacterial communities of healthy humans. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support ecological communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we empower future studies to  investigate how these community structures are influenced by disease state and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets that explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The twin and diet studies utilized multiple displacement amplification methods in their library preparations. <!--this is weird here, I would just bring it up in in the discussion of these studies or when you talk about the caveats of the overall approach/interpretation, cautioning about the fact that MDA was used in these two studies, and it will bias the representation (which nodes exist) and relative node quantification (node/cluster size) --> These datasets were selected because they included virome samples subjected to virus-like particle (VLP) purification. To this end, they employed combinations of filtration, chloroform/DNase treatment, and cesium chloride gradients to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei].

@@

# Discussion
Foundational work has provided a baseline understanding of the human microbiome by characterizing bacterial and viral diversity across the human body [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]<!--I think it ok to reframe/reset the discussion to focus on human-->. Here, we offer a novel approach to describe how phage-bacteria networks differ throughout the human body to provide a baseline for future studies of how and why microbiome networks differ in disease states. We developed and implemented a network-based analytical approach to evaluate the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity alone. This enabled the application of network theory to provide a new perspective on complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact through mechanisms such as horizontal gene transfer, modulated bacterial gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggested that microbiome network properties, such as stability <!--this is a network property -->, the potential for horizontal gene transfer between members, co-evolution, etc.<!--but these are not network properties...so this sentence doesn't make sense. But do you mean the "potential for community interactions inferred from network properties" were personal? So could change this sentence to: "These findings suggested that the community properties inferred from microbiome interaction network structures, such as stability, the potential for horizontal gene transfer between members, co-evolution of populations, etc., were person-specific. These properties may be impacted by personal factors ranging from the body's immune system to external environmental conditions, such as climate and diet."-->, were personal and may be impacted by personal factors ranging from the body's immune system to external environmental conditions, such as climate and diet.

The ability of environmental conditions to shape gut and skin microbiome interaction network structure was further supported by our finding that diet and skin location were associated with altered network structures. <!-- This is a more appropriate leader sentence for the next two anecdotes.--> We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, our findings provided evidence that high-fat diets were less connected than low-fat diets, and high-fat diets therefore may lead <!-- another spit infinitive--> to less stable communities with a decreased ability for microbes to directly influence one another. We supported this finding with the observation that obesity may have been associated with decreased network connectivity. Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their infectious interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, the skin environment influenced the microbiome interaction network structure, as well. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure (the exposed and intermittently exposed sites) were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Together these data suggested that body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes. This points to a link between microbiome and environmental stability, and warrants further investigation.

While these findings take us an important step closer to understanding the microbiome through interspecies relationships, there are caveats to and considerations regarding the approach. First, <!--"although our infection classification model is advantageous over existing models," ...this was not tested, so wouldn't say it, but it is also not needed--> as with most classification models, the infection classification model developed and applied is only as good as its training set--in this case, the collection of experimentally-verified positive and negative infection data, where genomes of all members are fully sequenced. Large-scale experimental screens for phage and bacteria infectious interactions that report high-confidence negative interactions (i.e., no infection) will provide more robust model training and improved model performance. Furthermore, just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to improve performance.

Second, although our analyses offer an informative proof of concept, this work was done retrospectively and relied on existing data up to seven years old. These archived datasets were limited by the technology and costs of the time <!--the MDA issue falls under this, in my mind, as issue of using old datsets. So integrated it here, also to bring down the focus a bit/not give it it's own paragraph-->. This resulted in small sequencing effort (as compared to today's dataset sizes) and thus datasets that were suboptimally powered for the statistical analyses we strive for today. Further, two studies, the diet and twin studies, relied on multiple displacement amplification (MDA) in their library preparations--an approach used to overcome the large nucleic acids requirements, especially of older sequencing library generation protocols. MDA results in significant biases in microbobial community composition <!--Yilmaz, S., Allgaier, M. & Hugenholtz, P., 2010, Multiple displacement amplification compromises quantitative analysis of metagenomes, Nat Methods, 7(12), pp. 943-4-->, as well as toward ssDNA viral genomes [@Kim:2008to; @Kim:2011hp], thus rendering the resulting microbial and viral metagenomes non-quantitative. <!--because Yilmaz et al., show that the bias depends on the initial community structure, I think this means that an MDA bias would also compromise your ability to compare samples even within studies, so I removed this: "However, by comparing samples within studies, which were all treated the same, we were able to provide the fairest possible comparison for drawing conclusions."--> Future work that employs larger sequence datasets and that avoids the use of MDA will build on and validate our findings, as well as inform the design and interpretation of further studies.

Finally, the networks in this study were built using operational genomic units, which represent groups of highly similar bacteria or phage genomes or genome fragments as clustered sub-populations. This operationally defined approach allows us to study whole community networks, but limits our ability to make conclusions about interactions among specific phage or bacterial species or populations. Although this approach lacks the resolution for drawing conclusions about specific bacteria or phages, future work could begin addressing these questions through improved binning methods and deeper metagenomic shotgun sequencing, as well as an improved understanding of what defines ecologically and evolutionarily cohesive units for both phage and bacteria, e.g., Polz et al., 2006 <!--Polz, M.F., Hunt, D.E., Preheim, S.P. & Weinreich, D.M., 2006, Patterns and mechanisms of genetic and phenotypic differentiation in marine microbes, Philos Trans R Soc Lond B Biol Sci, 361(1475), pp. 2009-21. -->

Together our work takes an initial step towards defining bacteria-virus interaction profiles as a characteristic of human-associated microbial communities. This approach revealed the impacts different human environments (e.g., the skin and gut) can have on microbiome connectivity. By focusing on relationships between bacterial and viral communities, they are studied as the interacting cohorts they are, rather than as independent entities. While our developed bacteria-phage interaction framework is a novel conceptual advance, the microbiome also consists of archaea and small eukaryotes, such as fungi and *Demodex* mites <!--reference?-->. Further, the components of the microbiome can interact with human immune cells and other non-microbial community members <!--reference?-->. Future work will build from our approach and include these additional community members and their diverse interactions and relationships (e.g., beyond phage-bacteria). This will result in a more robust network and a more holistic understanding of the human-associated microbiome and the evolutionary and ecological processes that drive its assembly and function <!--there were a lot of redunant thougths here I streamlined. Also you'll see I merged the final paragraphs to make the shared forward looking elements stronger.-->

@@

## Data Availability
All associated source code is available on GitHub at the following repository <!--don't know if you want to change this to reflect journal change?-->:

@@

## Bacterial OGU Identification
The longest contig from each OGU was used as its representative sequence for identification. Representative sequences were aligned to the European Nucleotide Archive (ENA) bacterial reference database using the blastn algorithm (e-value < $10^{-5}$). 79% of the bacterial OGUs were able to be identified. <!--what was this ID used for? i don't recall any discussion of inferred cluster taxonomic identity-->

## Phage OGU Identification
To confirm a lack of phage sequences in the bacterial OGU dataset, we performed a relatively strict <!--indeed, this is relative. can you relate it? What is the distribution of evalues or bitscores from this blastp? any anecdotes that it was a useful cut-off?--> blast nucleotide alignment of the bacterial OGU representative sequences using an e-value < $10^{-25}$. We used a stricter threshold because we know there are genomic similarities between bacteria and phage OGUs from the interactive model, but we were interested in contigs with high enough similarity to references that they may indeed be from phages. 2% of the OGUs had nucleotide similarities to known bacteriophage genomes, although the alignments were short (a couple of kb) and represented a small fraction of the alignment sequences. <!-- shouldn't you also consider a relative length component? Meaning, contigs with only phage similarity along its entire length are more likely phage than contigs with only small portion similar to phage, while rest has similarity to microbes...thus more likley prophage and ok to stay classified as bacteria. Why not just run already created probabilitic models that use many criteria, like VirSorter? Then you get predictions of phage (cat 1-3) versus propahge (cat 4-6) and the confidence of the prediction. -->

>>>>

***Journal***: mSystems

@@

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying microbes and viruses independently, thereby reducing them to two separate communities. Such approaches are unable to capture how these microbial communities interact, such as through processes that maintain community stability or allow phage-host populations to co-evolve. We developed and implemented a network-based analytical approach to describe phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages could infect which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. This study quantified and contrasted the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to better understand system perturbations, such as disease states, through ecological networks.

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages) have been also associated with various diseases and perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. The viruses act in concert with their microbial hosts as a single ecological community [@Haerter:2014ii]. Viruses influence their living microbial host communities through processes including lysis, modulating host gene expression [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja], influencing evolutionary processes, such as horizontal gene transfer [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn] or antagonistic co-evolution [@Koskella:2014ds], and by altering ecosystem processes and elemental stoichiometry [@Jover:2014gq].

Previous human microbiome work has focused on bacterial and viral communities, but have reduced them to two separate communities by studying them independently  [@Norman:2015kb; @Manrique:2016dx; @Ly:2014ew; @Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of bacteria or phage can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. Relationship-based network approaches allow us to capture this interaction information. We therefore leveraged machine learning and graph theory techniques to characterize human bacterial and phage communities by their inferred relationships. By doing so, we provide a foundation for further studies of disease network dynamics and gain broader insights into human microbiome network diversity across different body sites.

Human bacterial-phage network diversity was established and explored by utilizing three published microbiome datasets that contained paired virus and whole community metagenomic sequence sets dominated by bacteria [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques, which limit the scale of the experiments and analyses [@Flores:2013hc]. This metagenomic interaction inference model built on previous models, which have utilized a variety of techniques including taxonomic-based co-occurrence models using linear relationships between bacteria and phages [@LimaMendez:2015hw], as well as nucleotide similarity (including CRISPR targeting) models using OGUs [@@Roux:2016cc] and whole genomes [@Edwards:2015iz]. Our approach uniquely included protein interaction data, and was validated on experimentally determined negative and positive interactions, i.e., "who does and does not infect whom." Through this approach, we were able to provide a basic understanding of the network dynamics associated with phage and bacterial communities on the human body. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support ecological communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we empower future studies to  investigate how these community structures are influenced by disease state and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets that explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The twin and diet studies utilized multiple displacement amplification methods in their library preparations. These datasets were selected because they included virome samples subjected to virus-like particle (VLP) purification. To this end, they employed combinations of filtration, chloroform/DNase treatment, and cesium chloride gradients to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei].

@@

# Discussion
Foundational work has provided a baseline understanding of the human microbiome by characterizing bacterial and viral diversity across the human body [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Here, we offer an approach to describe how phage-bacteria networks differ throughout the human body to provide a baseline for future studies of how and why microbiome networks differ in disease states. We developed and implemented a network-based analytical model to evaluate the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity alone. This enabled the application of network theory to provide a new perspective on complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact through mechanisms such as horizontal gene transfer, modulated bacterial gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggested that the community properties inferred from microbiome interaction network structures, such as stability, the potential for horizontal gene transfer between members, co-evolution of populations, etc., were person-specific. These properties may be impacted by personal factors ranging from the body's immune system to external environmental conditions, such as climate and diet.

The ability of environmental conditions to shape gut and skin microbiome interaction network structure was further supported by our finding that diet and skin location were associated with altered network structures. We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, our findings provided evidence that high-fat diets were less connected than low-fat diets, and high-fat diets therefore may lead to less stable communities with a decreased ability for microbes to directly influence one another. We supported this finding with the observation that obesity may have been associated with decreased network connectivity. Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their infectious interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, the skin environment influenced the microbiome interaction network structure as well. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure (the exposed and intermittently exposed sites) were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Together these data suggested that body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes. This points to a link between microbiome and environmental stability, and warrants further investigation.

While these findings take us an important step closer to understanding the microbiome through interspecies relationships, there are caveats to and considerations regarding the approach. First, as with most classification models, the infection classification model developed and applied is only as good as its training set -- in this case, the collection of experimentally-verified positive and negative infection data, where genomes of all members are fully sequenced. Large-scale experimental screens for phage and bacteria infectious interactions that report high-confidence negative interactions (i.e., no infection) will provide more robust model training and improved model performance. Furthermore, just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to improve performance.

Second, although our analyses offer an informative proof of concept, this work was done retrospectively and relied on existing data up to seven years old. These archived datasets were limited by the technology and costs of the time. This resulted in small sequencing effort (as compared to today's dataset sizes) and thus datasets that were sub-optimally powered for the statistical analyses we strive for today. Further, two studies, the diet and twin studies, relied on multiple displacement amplification (MDA) in their library preparations--an approach used to overcome the large nucleic acids requirements, especially of older sequencing library generation protocols. MDA results in significant biases in microbial community composition [@Yilmaz:2010jb], as well as toward ssDNA viral genomes [@Kim:2008to; @Kim:2011hp], thus rendering the resulting microbial and viral metagenomes non-quantitative. Future work that employs larger sequence datasets and that avoids the use of MDA will build on and validate our findings, as well as inform the design and interpretation of further studies.

Finally, the networks in this study were built using operational genomic units, which represent groups of highly similar bacteria or phage genomes or genome fragments as clustered sub-populations. This operationally defined approach allows us to study whole community networks, but limits our ability to make conclusions about interactions among specific phage or bacterial species or populations. Although this approach lacks the resolution for drawing conclusions about specific bacteria or phages, future work could begin addressing these questions through improved binning methods and deeper metagenomic shotgun sequencing, as well as an improved understanding of what defines ecologically and evolutionarily cohesive units for both phage and bacteria, e.g., Polz et al., 2006 [@Polz:2006fi].

Such work will include more sophisticated approaches to defining operational genomic units and their taxonomic underpinnings (e.g. OGUs defined as clustering at the genus or species level). Previous work has implemented more sophisticated clustering and validation approaches that allow for such refined operational genomic unit clustering, such as the phylogenomic analyses performed to cluster cyanophage isolate genomes into informative groups using shared gene content, average nucleotide identity of shared genes, and pairwise differences between genomes [@Gregory:2016cg]. Similar clustering definition and validation methods, both computational and experimental, have been implemented in some metagenomic sequencing studies [@Roux:2016cc; @Brum:2015iaa; @Deng:2014eb; @Minot:2012ed] and could offer yet another level of sophistication to our network-based analyses, especially by allowing us to make conclusions about the specific virus ad bacterial interactions within the networks.

Together our work takes an initial step towards defining bacteria-virus interaction profiles as a characteristic of human-associated microbial communities. This approach revealed the impacts different human environments (e.g., the skin and gut) can have on microbiome connectivity. By focusing on relationships between bacterial and viral communities, they are studied as the interacting cohorts they are, rather than as independent entities. While our developed bacteria-phage interaction framework is a novel conceptual advance, the microbiome also consists of archaea and small eukaryotes, including fungi and *Demodex* mites [@Hannigan:2013im; @Grice:2011gy]. Further, the components of the microbiome can interact with human immune cells and other non-microbial community members [@Round:2009bz]. Future work will build from our approach and include these additional community members and their diverse interactions and relationships (e.g., beyond phage-bacteria). This will result in a more robust network and a more holistic understanding of the human-associated microbiome and the evolutionary and ecological processes that drive its assembly and function.

@@

## Data Availability
All associated source code is available on GitHub at the following repository:

@@

## Phage OGU Identification
To confirm a lack of phage sequences in the bacterial OGU dataset, we performed blast nucleotide alignment of the bacterial OGU representative sequences using an e-value < $10^{-25}$, which was stricter than the $10^{-10}$ threshold used in the random forest model below. We used a stricter threshold because we know there are genomic similarities between bacteria and phage OGUs from the interactive model, but we were interested in contigs with high enough similarity to references that they may indeed be from phages. 2% of the OGUs had nucleotide similarities to known bacteriophage genomes, although the alignments were short (a couple of kb) and represented a small fraction of the alignment sequences.
